Gaps in stellar streams as a result of globular cluster fly-bys

### 2024-11-18 paper/main.tex
integrated paola's intro rewrite

\title{Palomar 5 Gaps}

\subtitle{Globular clusters as hole punchers}

\author{S. Ferrone
       \inst{1,2}
       \and
       P. Di Matteo\inst{2}
       \and
       M. Montuori\inst{1}
       \and
       A. Mastrobuono-Battisti
       }

@@

\abstract
  {ok}
 % aims heading (mandatory)
  {ok}
 % methods heading (mandatory)
  {ok}
 % results heading (mandatory)
  {ok}
 % conclusions heading (optional), leave it empty if necessary
  {done}

\maketitle
\section{Introduction}
  Over a century ago, \citet{1933AcHPh...6..110Z} postulated the existence of dark matter to explain the excessively large velocity dispersions of galaxies within their galaxy clusters. Since then, mounting evidence has been assembled to support this claim. \citet{1970IAUS...38..351F,1970ApJ...161..802F} elaborated on the dynamics of disk galaxies and briefly acknoledged in the appendix of his paper that the rotation curves of NGC~300 and M~33 cannot be fully explained by disk models and that there must be the same amount of undetected mass as luminous mass and beyond the extent of the galaxies to explain their rotation curves. Concurrently \citet{1970ApJ...159..379R} studied the rotation of Andromenda and observed it to be flat outside its effective radius---not decreasing as one would expect from a gravitational field from a centrally concentrated mass distribution.

  Eventually, astronomers returned to Zwicky's non-luminous matter to explain these fast-rotating disks, as discussed in \citet{1983SciAm.248f..96R}'s review. Beyond the galactic scale, dark matter is essential for explaining many cosmological phenomena. For instance, the Planck mission \citep{2020A&A...641A...1P} was designed to study the Cosmic Microwave Background (CMB), whose spectrum closely follows that of a black body with a peak temperature in the millimeter wavelength range, corresponding to approximately 2.725~K. However, there are slight temperature variations on the order of one part in ten thousand. Notably, the colder regions of the CMB temperature map correlate with whole-sky gravitational lensing maps \citep{2020A&A...641A...8P}--since the photons lost energy as they escaped the gravitational wells. The $\Lambda$ Cold Dark Matter ($\Lambda$-CDM) model accounts for these observations almost perfectly, proposing that dark matter exists and that its distribution originated from stochastic quantum fluctuations within the first second of the universe led to density variations. These variations propagated as the universe expanded, shaping the large-scale structure and distribution of galaxies we observe today.

@@

  Stellar streams, several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy. They are one of the many pieces of evidence supporting the hierarchical nature of the galaxy formation as predicted by the $\Lambda CDM$ model \citep{1991ApJ...379...52W}. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, unless in the closest vicinity to the object \citep{2007ApJ...659.1212M}. Numerical predictions of this phenomenon exist since the 1970s \citep[see, for example][]{1975AJ.....80..290K}. These predictions occured well before the first detections which were made by \citet{1995AJ....109.2553G} around some Galactic globular clusters. Interestingly, this work was a nearly contemporary discovery of the Sagittarius stellar stream by \citet{1994Natur.370..194I} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way. Subsequent studies\footnote{\citep[see][for more subsequent observation detections of tidal debris, i.e. globular cluster stars beyond the tidal radius.]{1997A&A...320..776L, 2000A&A...356..127T, 2000A&A...359..907L, 2001AAS...19910906S, 2003AJ....126..815L}} extended Grillmair's findings to other globular cluster streams, until the discovery made by \citet{2001ApJ...548L.165O,2002AAS...200.1001O, 2003AJ....126.2385O} of long and thin tails outside the Palomar~5 globular cluster, which were shown to be extended for more than $10^\circ$ on the sky. These early works stimulated a vigorous search for stellar streams in the Galaxy in the following years \citep{2006ApJ...643L..17G, 2006ApJ...637L..29B, 2006ApJ...637L..29B, 2009ApJ...693.1118G}. The review of \citet{2024arXiv240519410B} properly asserts that we have envolved into a new area of stellar stream astronomy Gaia astrometric mission \citep{2016A&A...595A...1G}.

  Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and spectroscopy. The development and deployment of the Streamfinder algorithmm, first presented in \citet{2018MNRAS.477.4063M} has led to the detection of about 90 stellar streams in the Milky Way \citet{2024ApJ...967...89I}. \citet{2023MNRAS.520.5225M} provides a summary of these streams and tabulates that about 20 are associated to modern day globular clusters. One of the interests in studying stellar streams, is that they can constrain the gravitational field of their host galaxies, particularly the Milky Way. Compared to the measurement of the HI rotation curve, stellar streams offer the opportunity to investigate the potential of the host galaxy over a wide range of distances, reaching the outermost regions of the halo. For example, \citet{2011MNRAS.417..198V} demonstrated how stellar streams can infer the mass and scale parameters of dark matter haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \citet{2018ApJ...867..101B} reviewed this concept from an information-theoretic point of view, identifying which orbits and configurations of stellar streams that yield the most information about the galactic potential. However, the use of single streams for constraining the Galactic potential has led, however, to some ambiguous and not converging results over time. For example, \citet{2010ApJ...718.1128L} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{2004MNRAS.351..643H, johnston05, 2005ApJ...619..807L}; \citet{2016ApJ...833...31B} concluded that the dark matter halo of our Galaxy is nearly spherical at the radii of the Palomar~5 and GD~1 streams. Nonetheless, a recent goal of stellar stream astronomy was realized by \citet{2024ApJ...967...89I} constructed the most accurate model of the Milky Way to date based on a set of about 20 streams.

  % \paola{the following sentences seem to me a bit out of the focus of the discussion. It is interesting to say that a number of streams do not have a GC counterpart, but I would'nt go too much into the details, bcs you risk to bring the reader into another topic, that of the detection/undetection of streams, that of the metal-poor streams, etc. I would reduce what follows, and start again from "One of the interests in studying stellar streams,.." }Thus, currently there are about 70 progenitor-less streams and about 145 stream-less clusteres. The solo-streams can be explained by a larger globular cluster population whose members have been disolved due to two body relaxation and tidal stripping. For instance, Z Wan 2020( can't find the citation???) reports that the Phenoix stream is dynamically consistent from a globular cluster yet far more metal poor than any cluster ever observed, galactic or otherwise. This implies that the stream is from a completely disolved globular cluster which comes from a population of clusters that have not survived until modern times.

@@

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by \citet{2008MNRAS.391.1685S}, $\Lambda$-CDM predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring and had a mass of $10^8$ solar masses, as described by \citet{2012Natur.481..341V}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{2021arXiv211101148A}]{2005JCAP...08..003G, wang20}.

  \citet{2002MNRAS.332..915I} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{2012ApJ...748...20C} expanded this idea, proposing that subhalos could create gaps in stellar streams during flyby encounters by passing closely enough to a segment of a stream and significantly changing the orbits of the closest stars. In this work, \citet{2012ApJ...748...20C} further explored this idea through a probabilistic model of encounter rates based on the expected subhalo distribution. \citet{2019ApJ...880...38B} provided potential observational evidence for such a gap, identifying an underdensity in the GD-1 stream that could not be explained by known objects, such as globular clusters. Interestingly enough the expected mass and size of the perturber causing the gap was inconsistent with the $\Lambda$CDM mass-size relationship presented in \cite{2017MNRAS.466.4974M}.

  Previous studies have predominantly focused on the creation of gaps in stellar streams due to dark matter subhalo passages, with some exceptions. \citet{2016MNRAS.463L..17A} studied the possibility of disc giant molecular clouds gausing gaps in cluster streams. \citet{2017NatAs...1..633P} explored the effects of a galactic bar causing gaps and perturbing other aspects of a stellar stream. To our knowledge, no study has yet been performed investigating the gravitational impact of globular clusters on stellar streams. Since clusters are dense stellar systems, distributed mainly in the Galactic halo, they too should in principle contribute to generating perturbation on stellar streams. If globular clusters can indeed create inhomogeneities and gaps in streams, this factor must be taken into account when using stellar streams to infer the population of dark matter subhaloes.

  In our study, we will fill this gap in the literature by studying the gravitational perturbations induced by the whole set of Galactic globular clusters on Palomar 5 stream. We choose Palomar~5 since it is one of the most extensively studied and well-constrained globular cluster streams, known for its ease of observation and very long tidal tails. We present a result from one of our simulations in Fig. \ref{fig:stream_on_sky}. The following sections detail our modeling of the stream, its gravitational interactions with the galaxy and other globular clusters, and the statistical analysis of perturbations impacting the stream.

>>>>

%\title{Gaps in stellar streams as result of globular cluster close passages: the case of Palomar 5}
%\title{The impact of globular clusters close passages on stellar streams: the case of Pal5}
%\title{Globular clusters fly-bys and gap creations in stellar streams.}
\title{Gaps in stellar streams as a result of globular cluster fly-bys}

\subtitle{The case of Palomar~5}

\author{S. Ferrone
       \inst{1,2}
         \and
       M. Montuori\inst{1}
       \and
       P. Di Matteo\inst{2}
       \and
       A. Mastrobuono-Battisti
       \and
       et al.
       }

@@

\abstract
  {Thin stellar streams, such as those resulting from the tidal disruption of globular clusters, have long been known and used as probes of the gravitational potential of our Galaxy, both its visible and dark contents. In particular, the presence of under-density regions, or gaps, along these streams is commonly interpreted as being due to the close passage of dark matter sub-haloes. }
 % aims heading (mandatory)
  {In this work, we investigate the perturbations induced on streams by the passage of dense stellar systems, such as globular clusters themselves, to test the possibility that they too may cause the formation of gaps. In particular, we focus on the study of the stream of Palomar 5, a well-known globular cluster in the Galactic halo, which has particularly long tidal tails.  }
 % methods heading (mandatory)
  {For this purpose, we used a particle-test code to simulate the dynamic evolution of Palomar~5 and its tidal tails when Pal~5 is subject to the gravitational action of the Galaxy and its whole system of globular clusters.}
 % results heading (mandatory)
  {Our study shows that the tails of Pal~5 are strongly perturbed by the close passage of other clusters, in particular of NGC~2808, NGC~7078, NGC~104, and that these perturbations induce the formation of gaps in the tails.}
 % conclusions heading (optional), leave it empty if necessary
  {This result suggests that baryonic dense stellar systems such as globular clusters are capable of causing gaps in streams, as already demonstrated in the case of galaxy bar-induced perturbations and giant molecular clouds, and reinforce the idea that the effect of these components should be fully addressed when interpreting the morphology and density structures of observed stellar streams.}

\maketitle
\section{Introduction}

  Over a century ago, \citet{1933AcHPh...6..110Z} postulated the existence of dark matter to explain the excessively large velocity dispersions of galaxies within their galaxy clusters. Since then, mounting evidence has been assembled to support this claim. \citet{1970IAUS...38..351F,1970ApJ...161..802F} elaborated on the dynamics of disk galaxies and briefly acknowledged in the appendix of his paper that the rotation curves of NGC~300 and M~33 cannot be fully explained by disk models and that there must be the same amount of mass or more that is undetected and beyond the extend of the galaxies to explain their kinematics. Concurrently \citet{1970ApJ...159..379R} studied the rotation of Andromeda and observed it to be flat---not decreasing as one would expect from a gravitational potential with more mass concentration in the center. Eventually, astronomers returned to Zwicky's non-luminous matter to explain these fast-rotating disks, as discussed in \citet{1983SciAm.248f..96R}'s review.

  Beyond the galactic scale, dark matter is essential for explaining many cosmological phenomena. For instance, the Planck mission \citep{2020A&A...641A...1P} was designed to study the Cosmic Microwave Background (CMB), whose spectrum closely follows that of a black body with a peak temperature in the millimeter wavelength range, corresponding to approximately 2.725~K. However, there are slight temperature variations on the order of one part in ten thousand. Notably, the colder regions of the CMB temperature map correlate with whole-sky gravitational lensing maps \citep{2020A&A...641A...8P}--since the photons lost energy as they escaped the gravitational wells. The $\Lambda$ Cold Dark Matter ($\Lambda$-CDM) model accounts for these observations almost perfectly, proposing that dark matter exists and that its distribution originated from stochastic quantum fluctuations within the first second of the universe led to density variations. These variations propagated as the universe expanded, shaping the large-scale structure and distribution of galaxies we observe today.

@@

  Stellar streams are several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, unless in the closest vicinity to the object \citep{2007ApJ...659.1212M}. Numerical predictions of this phenomenon exist since the 1970s \citep[see, for example][]{1975AJ.....80..290K}. These predictions occured well before the first detections which were made by \citet{1995AJ....109.2553G} around some Galactic globular clusters.

  Interestingly, this work was a nearly contemporary discovery of the Sagittarius stellar stream by \citet{1994Natur.370..194I} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way. Subsequent studies\footnote{\citep[see][for more subsequent observation detections of tidal debris, i.e. globular cluster stars beyond the tidal radius.]{1997A&A...320..776L, 2000A&A...356..127T, 2000A&A...359..907L, 2001AAS...19910906S, 2003AJ....126..815L}} extended Grillmair's findings to other globular cluster streams, until the discovery made by \citet{2001ApJ...548L.165O,2002AAS...200.1001O, 2003AJ....126.2385O} of long and thin tails outside the Palomar~5 globular cluster, which were shown to be extended for more than $10^\circ$ on the sky. The discovery of its prominent tails stimulated a vigorous and successful search in the following years. New streams were discovered, mostly taking advantage of SDSS data, but also Pan-STARRS and ATLAS  \citep{2006ApJ...643L..17G, 2006ApJ...637L..29B, 2009ApJ...693.1118G, 2012ApJ...760L...6B, 2013ApJ...769L..23G, 2014ApJ...790L..10G, 2015ApJ...812L..26G, 2014MNRAS.443L..84B, 2016MNRAS.463.1759B, 2017ApJ...847..119G, 2014MNRAS.442L..85K}.

  It is with the publication of the data from the Gaia astrometric mission \citep{2016A&A...595A...1G} that a new era for stellar stream astronomy has started  \citep{2024arXiv240519410B}. Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and, for the brightest stars, spectroscopy. The possibility given by Gaia to track stars with coherent movements, over the entirety of the sky, has led to the discovery of dozens of new streams, as Slidr, Sylgr, Ylgr, Fimbulthul, Svol, Gjoll, Leiptr, Phlegeton and Fjorm \citep{2018MNRAS.481.3442M,  2018ApJ...865...85I, 2019ApJ...872..152I}, this latter having an extension of more than 150 degrees in the sky, that is more than five times greater (in projection) than that of Pal~5, according to the most recent estimates provided by \citet{2020MNRAS.493.4978S}. The possibility to couple Gaia data with spectroscopic surveys, as for example $S^5$ \citep{2019MNRAS.490.3508L},  has extended the study of stellar streams beyond the quantification of their orbital properties, to a full chemical characterization \citep{2020AJ....160..181J, 2021ApJ...911..149L, 2022ApJ...928...30L, 2024MNRAS.529.2413U}.  About a hundred stellar streams are currently known in our Galaxy, about 20 of which are associated to known Galactic globular clusters  \citep[see][for a compilation]{2023MNRAS.520.5225M}.

  One of the interests in studying stellar streams is that they can constrain the gravitational field of their host galaxies, particularly the Milky Way. Compared to the measurement of the HI rotation curve, stellar streams offer the opportunity to investigate the potential of the host galaxy over a wide range of distances, reaching the outermost regions of the halo. For example, \citet{2011MNRAS.417..198V} demonstrated how stellar streams can infer the mass and scale parameters of dark matter haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \citet{2018ApJ...867..101B} reviewed this concept from an information-theoretic point of view, identifying which orbits and configurations of stellar streams yield the most information about the galactic potential. However, the use of single streams for constraining the potential led to some ambiguous and not converging results over time. For example, \citet{2010ApJ...718.1128L} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{2004MNRAS.351..643H, 2005ApJ...619..800J, 2005ApJ...619..807L}; \citet{2016ApJ...833...31B} concluded that the dark matter halo of our Galaxy is nearly spherical at the distances of the Palomar~5 and GD~1 streams. Probably only by using the set of known streams can the degeneracy of the solution be eliminated. This approach has been recently used by \citet{2024ApJ...967...89I} who constructed the most accurate model of the Milky Way.

@@

  Beyond the global (visible and dark) mass distribution, as we anticipated earlier in this Introduction, streams can be used also to infer the granularity of the dark matter, that is the mass and density of the sub-halos populating our Galaxy. According to simulations by \citet{2008MNRAS.391.1685S}, $\Lambda$-CDM predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring and had a mass of $10^8$ solar masses, as described by \citet{2012Natur.481..341V}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{2021arXiv211101148A}]{2005JCAP...08..003G, wang20}.

  \citet{2002MNRAS.332..915I} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{2012ApJ...748...20C} expanded this idea, proposing that subhalos could create gaps in stellar streams during flyby encounters, where a subhalo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars.  \citet{2012ApJ...760...75C, 2013ApJ...768..171C} applied this idea and looked for the presence of gaps in the known Palomar~5 and GD~1 streams, concluding that the density variations found in their streams were consistent with expectations from $\Lambda$~CDM models.  \citet{2019ApJ...880...38B} provided further observational evidence for this idea, identifying a gap and a spur in the GD-1 stream that they could not explain by known objects, such as globular clusters, and which they suggested was due to the close passage of a dark matter sub-halo. Interestingly enough, however, the recovered properties of this sub-halo (mass and size) were inconsistent with the $\Lambda$CDM mass-size relationship presented in \cite{2017MNRAS.466.4974M}. The above-mentioned works are only a few of a much more extensive literature that has explored the impact of dark matter sub-haloes in simulated streams \citep{2016ApJ...828L..10H, 2021MNRAS.507.1999H, 2021JCAP...10..043B, 2024arXiv240402953H, 2024arXiv241021174N}, or searched for their traces in observed streams \citep{2016MNRAS.460.2711T, 2017MNRAS.470...60E, 2020ApJ...889...70B, 2020ApJ...892L..37B}.

  In contrast, a more limited number of works have been interested in understanding whether other structures, e.g. baryonics, can cause variations in the density of streams and gaps that can be assimilated and confused with those produced by dark matter sub-haloes.  Among these works, it is worth mentioning the results of \citet{2017NatAs...1..633P} that suggests how the presence of the bar at the center of the Galaxy can perturb the characteristics of a stream such as Pal~5 and generate gaps along its tail. That the Galactic bar could have an influence on streams morphology was also discussed by \citet{2016MNRAS.460..497H} and \citet{2016ApJ...824..104P}, in the case of the Ophiuchus stream \citep{2014MNRAS.443L..84B}. Beside the Galactic bar, also giant molecular clouds can produce gaps in stellar streams, as shown by \citet{2016MNRAS.463L..17A}. All these works thus seem to indicate that baryonic structures can play an important role in tail morphology. In this context,  an extensive numerical study, specifically focused on modelling the tails of Palomar~5, under the influence of the Galactic bar, spiral arms, giant molecular clouds and globular clusters, has been realised by \citet{2019MNRAS.484.2009B}, who concluded that both the influence of the bar and that of the giant molecular clouds can leave imprints in Palomar~5 tidal tails similar to those left by dark matter sub-halos, while the effect of globular clusters was found to be negligible.  A few additional studies have specifically investigated the effect of globular clusters on stellar streams.  Beside the work by \citet{2017MNRAS.470...60E}, who concluded that density variations, and in particular the dips, they found in Palomar~5 tails could not be due to the passage of globular clusters (but not that this conclusion was not based on a numerical modelling of this effect), more recently \citet{2022ApJ...941..129D} have examined the possibility that the gaps found in the GD-1 stellar stream could be due to the close passage of globular clusters, concluding that this scenario is highly unlikely. These first works  seem thus to suggest that the impact of globular clusters on stellar streams is negligible, and this is somewhat surprising, especially for streams of clusters as Palomar~5, which are confined in the inner 20~kpc of the Galaxy, where many other globular clusters do also orbit.

  In the present study we wish to fill this gap in the literature by reinvestigating this issue and quantify the importance of the effects of passing globular clusters in the vicinity of streams, to understand whether these systems can also be effective - and how often - in altering the distribution of stars in the tails and producing under-dense regions or gaps. To this end, on the following pages we will present the results of simulations of the streams of Pal5, subject to the gravitational interaction with the set of 165 Galactic globular clusters for which positions and velocities are known to date, and for which orbits can therefore be reconstructed. We chose to simulate Pal~5 because it is a typical halo cluster with extended tails, and because it is a cluster for which the effect of baryonic structures on its stream has already been studied. As we will show, and in tension with previous claims, the close passage of other clusters with the stream of Pal~5 is not a rare event, and indeed, in the 50 simulations we ran, we find the formation of numerous gaps, for an average of 1.5 gaps per simulation, generated by 18 different clusters among the whole system of Galactic globular clusters.
